\chapter{CONCLUSIONS}
\label{chapter6}

This chapter summarizes the entire Thesis and presents the final conclusions drawn from the study. It integrates the key findings from previous chapters and highlights the overall contributions of the research. The purpose of this chapter is not simply to repeat earlier content but to provide a clear and concise interpretation of what has been achieved.

This chapter is organized as follows: Section 6.1 provides a general conclusion, Section 6.2 highlights key findings, Section 6.3 discusses limitations of the study, Section 6.4 presents practical implications, and Section 6.5 provides recommendations for future research.

% -------------------------------------------------
\section{General}

This section provides a brief summary of the study, including the research objectives, methodology, and overall outcomes. It should highlight how the research problem was addressed and the extent to which the objectives were achieved.

Rather than repeating the contents of previous chapters, this section should reflect on the effectiveness of the methodology and discuss how well the results answered the research questions.

% -------------------------------------------------
\section{Key Findings}

This section presents the most important findings of the study. These findings should be derived from the results and discussion presented in previous chapters.

For example:
\begin{itemize}
    \item Identification of major trends or patterns in the data
    \item Achievement of the research objectives
    \item Key improvements or contributions introduced by the study
\end{itemize}

Clearly state the contribution of the research to the field. This may include theoretical advancement, methodological improvement, or practical application.

% -------------------------------------------------
\section{Limitation of the Study}

This section outlines the limitations of the study. These limitations may arise from constraints such as:

\begin{itemize}
    \item Limited data availability or quality
    \item Simplifying assumptions made during analysis
    \item Constraints in experimental setup or tools
\end{itemize}

Discuss how these limitations may have affected the results and how they can be addressed in future work.

% -------------------------------------------------
\section{Practical Implication}

This section discusses the practical significance of the research findings. Explain how the results can be applied in real-world scenarios.

Possible points to include:
\begin{itemize}
    \item Applications of the proposed approach
    \item Benefits for industry or academia
    \item Impact on future technological development
\end{itemize}

% -------------------------------------------------
\section{Recommendation for Future Study}

This section provides suggestions for further research based on the findings and limitations of the current study.

Future work may include:
\begin{itemize}
    \item Extending the study to larger datasets or different conditions
    \item Improving methodology or experimental design
    \item Exploring additional variables or techniques
\end{itemize}

This section demonstrates the potential for continued research and development in the area.

\vspace{1cm}

\noindent In conclusion, the study contributes to the understanding of the research problem and provides a foundation for future work. A strong final statement can be included to emphasize the importance of the study and its broader impact.