\chapter{Literature Review}
\label{ch:literature_review}

There exists a substantial amount of work on hallucination detection and mitigation in language models, but these two concerns are seldom addressed in a single unified system. This chapter reviews the literature relevant to each method family. We first discuss the phenomenon of hallucination itself (Section~\ref{sec:lit_hallucination}), before covering black-box and reference-based detectors (Section~\ref{sec:lit_blackbox}), internal-state probing approaches (Section~\ref{sec:lit_probing}), mechanistic work on FFN layers which motivates our IAV signal (Section~\ref{sec:lit_ffn}), closed-loop and self-corrective systems (Section~\ref{sec:lit_closedloop}), and finally cross-architecture generalization studies (Section~\ref{sec:lit_crossarch}).

%% -------------------------------------------------
\section{Hallucination in Retrieval-Augmented Generation}
\label{sec:lit_hallucination}

The term ``hallucination'' in the context of language models refers to output that sounds fluent and confident but is factually unsupported, unverifiable, or outright contradictory to the provided source material. It is not a niche problem, having been identified as one of the most pressing reliability issues for modern LLMs in multiple survey papers~\cite{ji2023hallucination,huang2023survey,rawte2023survey}. In RAG specifically, entity-based knowledge conflicts between retrieved evidence and parametric memory can trigger hallucination even when the retriever supplies relevant passages~\cite{longpre2021entity}.

The taxonomy that recurs across these surveys draws a useful distinction. \emph{Intrinsic} hallucination occurs when the generated text directly contradicts the given source,the model says one thing while the evidence says another. \emph{Extrinsic} hallucination is subtler: the model introduces claims that are plausible and might even be true, but that go beyond what the available evidence actually supports. Both types are dangerous in a RAG setting, because the user typically assumes that a retrieval-grounded system is trustworthy.

%% -------------------------------------------------
\section{Black-Box, Statistical, and Reference-Based Detection}
\label{sec:lit_blackbox}

One natural approach for catching hallucinations is to treat the model as a black box and look for inconsistencies in its output. An example of this approach is SelfCheckGPT~\cite{manakul2023selfcheckgpt} which produces multiple answers for the same query and verifies if the factual information is consistent across samples. The appeal is clear: no need to access model weights or internal states. But there's a catch. If the model is wrong with high confidence systematically, all sampled responses will have the same fabrication and the method will not flag it. In addition, the generation of multiple responses per query comes with the inference cost, which is impractical for real-time applications.

Reference-based methods are a different approach. Instead of sampling multiple times, they directly compare the generated text with the retrieved evidence. For instance, RefChecker~\cite{hu2024refchecker} breaks down the model output into fine-grained claim triplets and checks each triplet against the reference documents. This can be quite effective for post hoc verification – but requires a separate verification pipeline, adds latency, and does not use the generating model’s own internal computation as an intervention signal.

Collectively, these families of methods provide useful baselines and important conceptual contributions. But none of them is quite appropriate for the setting we study here. They either operate \emph{after} generation (too late for closed-loop control), need repeated sampling (too expensive), or depend on external checkers (adding complexity and latency). What we need is a signal from within the generating model itself, something cheap, real-time and actionable.

%% -------------------------------------------------
\section{Internal-State Probing for Hallucination Detection}
\label{sec:lit_probing}

The idea behind internal-state probing is simple: if a transformer is producing hallucinated text, maybe its hidden representations already “know” something is wrong, even if the surface text looks okay. Several groups of researchers have followed this intuition.

Amongst the first to show this convincingly were Azaria and Mitchell~\cite{azaria2023saplma}. They trained simple classifiers on the hidden activations of language models, and showed that these classifiers could distinguish true from false statements at roughly \textbf{71--83\% accuracy} depending on the base model. The implication is obvious: internal representations contain information relevant to truthfulness that is not fully reflected in output probabilities. However, their SAPLMA approach is a pure detector, which produces a score but does nothing with it. There's no feedback loop from that score to any corrective action.

INSIDE (also called EigenScore) extends the internal-state idea by examining the eigenspectrum of hidden representations over multiple sampled outputs~\cite{chen2024inside}. Surface-level metrics miss uncertainty that can be revealed by the covariance structure of these representations. That's a good method, but it shares the same weakness as SelfCheckGPT: it still needs multiple generations which is expensive for online use.

MIND~\cite{su2024mind} is most related to our work, which performs \emph{real-time} hallucination detection based on internal states without manual annotation for training. This supports the more general claim that internal representations can be useful at live inference, not just at post-hoc analysis. However, as most detection-based methods, MIND does not employ the multi-action controller we develop here.

A consistent finding in the literature is that useful signals for hallucination detection can be found in the internal activations. Hidden states, attention patterns, FFN activity, uncertainty measures, all of them have provided useful information. But the missing piece is the connection to action. Detection of hallucination is useful, but it becomes \emph{practically} useful only if detection causes some change in system behavior. That’s the gap we’re trying to close.

%% -------------------------------------------------
\section{FFN Layers as Factual Memory and Motivation for IAV Probing}
\label{sec:lit_ffn}

Why look inside the feed-forward network specifically? The motivation comes from a fascinating line of mechanistic interpretability research.

According to Geva et al.~\cite{geva2021ffn}, Transformer FFN layers are similar to key-value memories. The “ keys ” are textual patterns that trigger the input . The “ values ” are the contributions that are written back into the residual stream , which determines what tokens the model will output next . They also noticed a depth gradient with lower layers encoding surface patterns and upper layers encoding more semantic and factual information. Further empirical work has examined how these stored associations can be updated and modified~\cite{qiu2024updating}.

This has a direct consequence on our work. If FFN layers store and express factual associations, then the intermediate activation \emph{inside} the FFN, before the down-projection collapses it back into the residual stream, should contain a particularly rich signal about what parametric knowledge the model is injecting. This is exactly what our IAV (Intermediate Activation Vector) captures.

This gives us two complementary views: the residual-stream signal (CEV) and the FFN-interior signal (IAV). CEV discusses context integration, how well the model has integrated the retrieved evidence with the query and the text generated so far. IAV tells us about parametric injection - what the feed-forward network is contributing from its stored knowledge. Hallucination is possible if these two signals are not in agreement or if the IAV displays abnormal patterns.

%% -------------------------------------------------
\section{Closed-Loop and Self-Corrective Generation}
\label{sec:lit_closedloop}

The idea of making RAG systems self-corrective is not entirely new, though existing approaches differ substantially from ours in mechanism and scope.

Self-RAG~\cite{asai2024selfrag} trains a language model to generate special ``reflection tokens'' that critique its own output and decide whether additional retrieval is needed. The model learns to retrieve, generate, and self-evaluate all within a single framework. It is an elegant design, but it requires training the model with reflection-token supervision, which means modifying the backbone. Our approach, by contrast, works with completely frozen models and adds the control layer externally.

CRAG~\cite{yan2024crag} takes a different tack. It adds a lightweight evaluator that assesses whether the retrieved documents are actually good enough to answer the query. If not, it triggers corrective retrieval—potentially even web search. CRAG directly addresses the ``bad retrieval'' failure mode, and it does so effectively. But its intervention signal is about \emph{retrieval quality}, not about what the generator is actually doing internally. Our system monitors the generator itself.

Abstention—the ability to simply say ``I don't know''—is another important piece of the puzzle. Wen et al.~\cite{wen2024abstention} survey this capability in LLMs and argue that knowing when to refuse is itself a safety mechanism. We agree, and ABSTAIN is one of the four actions available to our controller.

To the best of our knowledge, no prior work has combined all of the following elements in a single system: hallucination detection from the generating model's own dual CEV/IAV activations, a multi-action inference-time controller (accept/regenerate/re-retrieve/abstain), and operation on completely frozen backbones. Each individual ingredient exists in the literature; the combination and the closed-loop integration are what we contribute.

%% -------------------------------------------------
\section{Cross-Architecture Generalization}
\label{sec:lit_crossarch}

A persistent concern with activation-probing methods is whether they generalize. If you train a probe on one model's hidden states, does the approach transfer to a different model? Or are the learned patterns so architecture-specific that you need to start from scratch each time?

The evidence is mixed. HalluRAG~\cite{frohling2024hallurag} studied internal-state hallucination detection in a closed-domain RAG setting and found that MLP classifiers could reach up to \textbf{75\% test accuracy} but with limited generalizability across different experimental configurations. This is not surprising given how different models produce very different activation distributions, but it does underscore the need to test on more than one backbone.

CLAP (Cross-Layer Attention Probing)~\cite{zhao2024clap} goes further by evaluating fine-grained hallucination detection across \textbf{five LLMs and three task types}. Their results confirm that activation-based signals do transfer to some degree, but the strength and reliability of that transfer varies considerably across models.

Our evaluation is designed with this concern front and center. We test the identical CEV/IAV probing framework and controller design on three decoder-only transformers—Mistral-7B-Instruct-v0.2, LLaMA-3-8B-Instruct, and Qwen3-8B which differ in their tokenizers, training data, layer counts, and FFN intermediate dimensions. The framework stays the same; the probe weights and thresholds are calibrated per backbone. This gives us a controlled test of reusability: the architecture is shared, but each deployment requires backbone-specific training. That turns out to be an honest and practical characterization of where internal-state probing stands today.